Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-05-24 appendix.tex
had to add edge case per vu

%Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
Consider the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$; we restrict our attention to $x_1,y_1,x_2,y_2\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
Recall the divergence
\[
\shannonDivergence_{\tau_1,t^*}(\tau_1,\tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}).
\]
By Gibbs inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$.
Referring to Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
-(x^*)^2 = -x_1x_2+y_1w(x_1-x_2)
\]
and for $\{1\}$ and $\{1,2\}$ yields
\[
-(x^*)^2 = -x_1x_2+y_2w(x_2-x_1).
\]
The difference of these two equations simplifies to
\[
0 = (y_1w+y_2w)(x_1-x_2)
\]
and, since $y_1w+y_2w > 0$ by assumption, we have $x_1=x_2$.
This along with the previous equalities imply $x_1=x_2=x^*$.
We use the same process to show that, at the maximum, $y_1=y_2=y^*$, and thus that the divergence term attains a unique maximum over $x_1,y_1,x_2,y_2\in(0,1)$.

We now focus on the partial log-likelihood term.
Recall the partial log-likelihood takes the form
\[
\tilde{\ell}_{\tau_1,t^*}(\tau_1, \tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau_1, \tilde{t}).
\]
The partial log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
\[
\tilde{\ell}_{\tau_1,t^*}(\tau_1, \tilde{t}) = \sum_{j=1}^\nSiteSplits c_j\cdot\left(\log h_{j,1}(x_1)+\log h_{j,2}(y_1)+\log h_{j,3}(x_2)+\log h_{j,4}(y_2)+\log h_{j,5}(w)\right).
\]
Due to this additive form, any off-diagonal terms of the Hessian for this function will be zero, so it remains to show that the diagonal terms are nonpositive.
Without loss of generality we focus on the variable $x_1$ and the partial log-likelihood proportional to
\[
\tilde{\ell}(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,1}(x_1).
\]
Doing calculation as in Figure~\ref{fig:example_likelihoods} with $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$, we see each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, will be
\[
h_{j,1}(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
\]
for $e_j\in\{0,1\}$, which, simplifying, results in
\begin{align}
\tilde{\ell}(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1).
\end{align}
This has a second derivative
\[
\tilde{\ell}''(x_1) = -\left(\frac{\sum_j c_je_j}{(1+x_1)^2}+\frac{1-\sum_j c_je_j}{(1-x_1)^2}\right).
\]
Again, by assumption $x_1,y_1,x_2,y_2\in(0,1)$, and we need only $0 \le \sum_j c_je_j \le 1$ so that the diagonal term of the Hessian will be nonpositive.
Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\tilde{\ell}''(x_1) \le 0$.
Applying similar arguments to the other variables, the Hessian for the partial log-likelihood will have nonpositive diagonal terms and off-diagonal terms equal to zero.
Thus, the Hessian of the sum of the divergence and the partial log-likelihood will be negative definite, as the Hessian for the divergence is negative definite, and there exists a unique maximum to the likelihood over $x_1,x_2,y_1,y_2\in(0,1)$.
Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

>>>>

%Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
Consider the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$; we restrict our attention to $x_1,y_1,x_2,y_2\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
Additionally, we exclude the infinite branch length case of $w=0$.
Recall the divergence
\[
\shannonDivergence_{\tau_1,t^*}(\tau_1,\tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}).
\]
By Gibbs inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$.
Referring to Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
-(x^*)^2 = -x_1x_2+y_1w(x_1-x_2)
\]
and for $\{1\}$ and $\{1,2\}$ yields
\[
-(x^*)^2 = -x_1x_2+y_2w(x_2-x_1).
\]
The difference of these two equations simplifies to
\[
0 = (y_1w+y_2w)(x_1-x_2)
\]
and, since $y_1w+y_2w > 0$ by assumption, we have $x_1=x_2$.
This along with the previous equalities imply $x_1=x_2=x^*$.
We use the same process to show that, at the maximum, $y_1=y_2=y^*$, and thus that the divergence term attains a unique maximum over $x_1,y_1,x_2,y_2\in(0,1)$.

We now focus on the partial log-likelihood term.
Recall the partial log-likelihood takes the form
\[
\tilde{\ell}_{\tau_1,t^*}(\tau_1, \tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau_1, \tilde{t}).
\]
The partial log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
\[
\tilde{\ell}_{\tau_1,t^*}(\tau_1, \tilde{t}) = \sum_{j=1}^\nSiteSplits c_j\cdot\left(\log h_{j,1}(x_1)+\log h_{j,2}(y_1)+\log h_{j,3}(x_2)+\log h_{j,4}(y_2)+\log h_{j,5}(w)\right).
\]
Due to this additive form, any off-diagonal terms of the Hessian for this function will be zero, so it remains to show that the diagonal terms are nonpositive.
Without loss of generality we focus on the variable $x_1$ and the partial log-likelihood proportional to
\[
\tilde{\ell}(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,1}(x_1).
\]
Doing calculation as in Figure~\ref{fig:example_likelihoods} with $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$, we see each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, will be
\[
h_{j,1}(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
\]
for $e_j\in\{0,1\}$, which, simplifying, results in
\begin{align}
\tilde{\ell}(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1).
\end{align}
Assume for now the functional form of the partial likelihood is fixed---in other words, each site split has a maximum at a single ancestral state split.
This has a second derivative
\[
\tilde{\ell}''(x_1) = -\left(\frac{\sum_j c_je_j}{(1+x_1)^2}+\frac{1-\sum_j c_je_j}{(1-x_1)^2}\right).
\]
Again, by assumption $x_1,y_1,x_2,y_2\in(0,1)$, and we need only $0 \le \sum_j c_je_j \le 1$ so that the diagonal term of the Hessian will be nonpositive.
Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\tilde{\ell}''(x_1) \le 0$.
Applying similar arguments to the other variables, the Hessian for the partial log-likelihood will have nonpositive diagonal terms and off-diagonal terms equal to zero.
Thus, the Hessian of the sum of the divergence and the partial log-likelihood will be negative definite, as the Hessian for the divergence is negative definite, provided there is a single maximal ancestral state split for each site split.

We now consider the case where there may be more than one ancestral state split that maximizes the partial likelihood for a site split.
For site split $\{1\}$, the four possible forms of the partial likelihood are
\[
(1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2),
\]
\[
(1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2),
\]
\[
(1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2),
\]
\[
(1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2).
\]
We ignore the second function as it will always be less than the first except for the case $y_1=y_2=w=0$.
Comparing the remaining functions reveals that there will be an ambiguity in the form of the likelihood only when either $y_1=y_2=0$, $x_1=x_2=0$, or $w=0$.
We reach the same conclusions doing these comparisons for site splits $\emptyset$, $\{2\}$, $\{3\}$, and $\{1,2,3\}$.
The remaining site splits $\{1,2\}$, $\{1,3\}$, and $\{2,3\}$ require the following argument.
For site split $\{1,2\}$, the four possible forms of the partial likelihood are
\[
(1-x_1)(1+x_2)(1+w)(1-y_1)(1+y_2),
\]
\[
(1-x_1)(1-x_2)(1-w)(1+y_1)(1-y_2),
\]
\[
(1+x_1)(1-x_2)(1-w)(1-y_1)(1+y_2),
\]
\[
(1+x_1)(1-x_2)(1+w)(1+y_1)(1-y_2).
\]
We ignore cases where either $y_1=y_2=0$, $x_1=x_2=0$, or $w=0$, and see that an ambiguity in this maximization will occur when
\[
(1-x_1)(1+x_2)(1-y_1)(1+y_2) = (1+x_1)(1-x_2)(1+y_1)(1-y_2).
\]
Note that this can happen when $x_1=y_2=u_1$ and $x_2=y_1=u_2$ but $u_1 \neq u_2$.
However, when $x_1=x_2=u_1$ and $y_1=y_2=u_2$ the partial likelihood takes the same value.
To resolve this ambiguity, we look to the divergence quanitity and recall it will be maximized when $x_1=x_2$ and $y_1=y_2$.
Thus, the full joint likelihood will obtain a unique maximum, and this maximum will occur when $x_1=x_2$ and $y_1=y_2$.

Given the above, there exists a unique maximum to the likelihood over $x_1,y_2,x_1,y_2\in(0,1)$ with $w\neq 0$, and, with the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.
